A Study of Institutional Spending on Open Access Publication Fees in Germany

### 2016-06-10 article.Rmd
some minor fixes

```{r results='hide'}
tt <- read.csv("data/aggregates_fees_paid.csv", header = TRUE, sep =",")
colnames(tt) <- c("Cost data-set", "Journal Type", "Articles funded", "Total costs in €", "Mean")
table_2 <- tt
#table_2 <- pander::pandoc.table(tt,split.cells = 5 , big.mark = ",", caption = 'Comparision of cost data per period and journal type (in €)', justify = 'lrrrr')
```

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 2 shows the top ten publishers in terms of the number of financially supported articles. These publishers represented `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of all articles included in our data-set. In total, payments were made to `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. Comparing these data with data from the UK, we observed that a greater share of total spending was directed to some open access publishers. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI, Copernicus, and Hindawi.

**Table 2: Publication fees paid per publisher (in €)**

@@

Prices also varied within journals. Table 3 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals, based on the number of supported articles. We normalised PLOS journal titles because the name change from "PLoS" to "PLOS" was only partially represented in the Crossref metadata at the time of our study. Articles published in the top ten journals represented `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall article volume. The multidisciplinary journal PLOS ONE ranked highest. In addition, the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline, were also well represented in our study. In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and takes the submission's file format into consideration.

@@

Re-using DOIs to gather bibliographic metadata from Crossref is a promising approach to addressing data curation issues while unifying UK spending data [@Neylon_2014; @woodward_2014]. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing information about publisher and journal titles for 99 % of all articles included in the Open APC data-set. Making use of metadata from Crossref, therefore, reduces the need for extensive validation of bibliographic records as long as the DOIs are made available in the cost data. Beyond identifying publishers and journals, mandatory reporting of DOIs in the spending data can also increase the use of such data to study other aspects of APC-based open access journals. For instance, impact analyses in the field of altmetrics make heavy use of DOIs as well [@Haustein_2016]. Drawing on Crossref has the potential to increase the comparability of cost data to prepare for future negotiations with publishers regarding open access agreements because Crossref's metadata represent current developments in academic publishing in terms of ongoing mergers of publishing houses. In addition to these practical benefits, future comparative studies of publication fee spending using data at the article level can also benefit from such an approach.

This study is limited in some respects. First, we cannot assess whether publishers and journals granted publication fee discounts. The Open APC initiative uses a minimal data scheme to encourage self-reporting, and it therefore does not track this type of information. However, large price variations suggest that different pricing levels and pricing schemes are in place, as previously observed [@Solomon_2012; @Pinfield_2015]. Adding to this complexity, it is likely that some institutions paid only part of the publication fee. Consider, for instance, the journal Nature Communication: Charges reported to the Open APC initiative ranged between € 2000, the DFG price cap, and € 4,403 for a journal that can be categorised as a pricy, high-impact journal according to @Solomon_2012. Although making such payments from several budgets is a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro-rata payments were not made transparent in the Open APC data, leading to a possibly flawed representation of publication fee spending in Germany. In another case, one university included its contributions to the SCOAP^3^ consortia and presumably divided the sum by the number of articles published by their authors in SCOAP^3^-covered journals.[^26] This approach is arbitrary, because averages for an institution can be determined only after the end of a full 3-year funding cycle. Other factors affecting price variations are exchange rates and different tax rates for some organisations in Germany. For instance, the Max Planck Society has a limited input tax reduction. The refund of input value-added tax for publication fees is 20 %. To increase the transparency of publication fee spending, @Pinfield_2015 suggested disclosing tax rates and payment currencies in future cost data-sets. In addition, the Open APC data-set does not track funding sources; thus, we could not determine which funders co-financed publication fees.

@@

One possible explanation for why traditional publishers are less well represented in our study is the lack of cost information about hybrid open access journals. In fact, 99 % of all articles that German universities and research organisations supported financially were published in fully open access journals. This result presumably reflects the DFG funding policy that opposes paying for open access articles published in hybrid journals. However, while reviewing self-reported cost data from Austria and the UK, where hybrid open access journals are generally supported, we observed a much higher share of payments for articles in hybrid open access journals. Because publication fee spending is fragmented and often lacks transparency, it remains open to speculation whether authors affiliated with German universities and research organisations avoid opting for open access when publishing in hybrid journals or whether they simply use other budgets that are not covered by the Open APC initiative.

@@

Our study revealed the size and extent of spending on open access journals using publication fees in Germany. According to self-reported cost data from the Open APC initiative, this type of support from German universities and research institutions has grown over the years. Comparing these expenditure with those from Austria and the UK, German open access funding is focussed primarily on fully open access journals, raising important questions about hybrid open access journals as a publication venue. Given our findings and the general discussion about funding policies addressing hybrid open access journals, questions about whether and to what extent science policy interventions and the availability of institutional support influence how researchers publish are of particular concern.

Using self-reported data and gathering publisher and journal information from Crossref, our study extends methods and improves data collection activities for researchers and practitioners alike, as well as contributing to a better understanding of the factors affecting the analysis of publication fees in open access publishing. In this regard, our research highlights large variations in the distribution of spending that need to be taken into consideration when studying payments on publications at the institutional level. We have also confirmed previous studies that showed large price variations across publishers and open access journals; these variations need to be better understood in future.

>>>>

```{r results='hide'}
tt <- read.csv("data/aggregates_fees_paid.csv", header = TRUE, sep =",")
colnames(tt) <- c("Cost data-set", "Journal Type", "Articles funded", "Total costs in €", "Mean")
table_2 <- tt
#table_2 <- pander::pandoc.table(tt,split.cells = 5 , big.mark = ",", caption = 'Comparision of cost data per period and journal type (in €)', justify = 'lrrrr')
```
**Table 2: Comparison of data by journal type and financial period**

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 3 shows the top ten publishers in terms of the number of financially supported articles. These publishers represented `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of all articles included in our data-set. In total, payments were made to `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. Comparing these data with data from the UK, we observed that a greater share of total spending was directed to some open access publishers. @Pinfield_2015, for instance, reported remarkably lower proportions for the open access publishers MPDI, Copernicus, and Hindawi.

**Table 3: Publication fees paid per publisher (in €)**

@@

Prices also varied within journals. Table 4 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals, based on the number of supported articles. We normalised PLOS journal titles because the name change from "PLoS" to "PLOS" was only partially represented in the Crossref metadata at the time of our study. Articles published in the top ten journals represented `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall article volume. The multidisciplinary journal PLOS ONE ranked highest. In addition, the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline, were also well represented in our study. In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and takes the submission's file format into consideration.

@@

Re-using DOIs to gather bibliographic metadata from Crossref is a promising approach to addressing data curation issues raised by UK initiatives [@Neylon_2014; @woodward_2014]. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing information about publisher and journal titles for 99 % of all articles included in the Open APC data-set. Making use of metadata from Crossref, therefore, reduces the need for extensive validation of bibliographic records as long as the DOIs are made available in the cost data. Beyond identifying publishers and journals, mandatory reporting of DOIs in the spending data can also increase the use of such data to study other aspects of APC-based open access journals. For instance, impact analyses in the field of altmetrics make heavy use of DOIs as well [@Haustein_2016]. Drawing on Crossref has the potential to increase the comparability of cost data to prepare for future negotiations with publishers regarding open access agreements because Crossref's metadata represent current developments in academic publishing in terms of ongoing mergers of publishing houses. In addition to these practical benefits, future comparative studies of publication fee spending using data at the article level can also benefit from such an approach.

This study is limited in some respects. First, we cannot assess whether publishers and journals granted publication fee discounts. The Open APC initiative uses a minimal data scheme to encourage self-reporting, and it therefore does not track this type of information. However, large price variations suggest that different pricing levels and pricing schemes are in place, as previously observed [@Solomon_2012; @Pinfield_2015]. Adding to this complexity, it is likely that some institutions paid only part of the publication fee. Consider, for instance, the journal Nature Communication, a journal that can be categorised as a pricy, high-impact journal according to @Solomon_2012: Charges reported to the Open APC initiative ranged between € 2000, the DFG price cap, and € 4,403. Although making such payments from several budgets is a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro-rata payments were not made transparent in the Open APC data, leading to a possibly flawed representation of publication fee spending in Germany. In another case, one university included its contributions to the SCOAP^3^ consortia and presumably divided the sum by the number of articles published by their authors in SCOAP^3^-covered journals.[^26] This approach is arbitrary, because averages for an institution can be determined only after the end of a full 3-year funding cycle. Other factors affecting price variations are exchange rates and different tax rates for some organisations in Germany. For instance, the Max Planck Society has a limited input tax reduction. The refund of input value-added tax for publication fees is 20 %. To increase the transparency of publication fee spending, @Pinfield_2015 suggested disclosing tax rates and payment currencies in future cost data-sets. In addition, the Open APC data-set does not track funding sources; thus, we could not determine which funders co-financed publication fees.

@@

One possible explanation for why traditional publishers are less well represented in our study is the lack of cost information about hybrid open access journals. In fact, 99 % of all articles that German universities and research organisations supported financially were published in fully open access journals. This result presumably reflects the DFG funding programme that excludes paying for open access articles published in hybrid journals. However, while reviewing self-reported cost data from Austria and the UK, where hybrid open access journals are generally supported, we observed a much higher share of payments for articles in hybrid open access journals. Because publication fee spending is fragmented and often lacks transparency, it remains open to speculation whether authors affiliated with German universities and research organisations avoid opting for open access when publishing in hybrid journals or whether they simply use other budgets that are not covered by the Open APC initiative.

@@

Our study revealed the size and extent of spending on open access journals using publication fees in Germany. According to self-reported cost data from the Open APC initiative, this type of support from German universities and research institutions has grown over the years. Comparing these expenditure with those from Austria and the UK, German open access funding is focussed primarily on fully open access journals, raising important questions about hybrid open access journals as a publication venue. Given our findings and the general discussion about funding policies addressing hybrid open access journals, questions about whether and to what extent science policies and the availability of institutional support influence how researchers publish are of particular concern.

Using self-reported data and gathering publisher and journal information from Crossref, our study extends methods and improves data collection activities for researchers and practitioners alike, as well as contributing to a better understanding of the factors affecting the analysis of publication fees in open access publishing. In this regard, our research highlights large variations in the distribution of spending that need to be taken into consideration when studying payments on publications at the institutional level. We have also confirmed the findings of other studies that showed large price variations across publishers and open access journals. This complex situation of fee-based open access publishing need to be better understood by both researchers and practitioners.
